### 2023-04-26 paper.Rmd
editing

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning.

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, two simulation studies show that k-plus anticlustering is well-suited to create groups having minimal differences according to multiple distribution characteristics, significantly outperforming traditional anticlustering criteria. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to create groups that have similar variances across the $P$ attributes. As a caveat, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. To reiterate, this condition has to be met for the original data $\mathbf{X}_{N \times P}$. Thus, only considering $\mathbf{Y}_{N \times P}$ in order to minimize discrepancies in variance may not yield satisfying results; instead, it has to be ensured that the group means of the original data are also similar between groups. Fortunately, minimizing discrepancy with regard to group means can be accomplished by standard k-means anticlustering, i.e., the maximization of $\mathit{SSE}$. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

## A multicriterion optimization model: K-plus anticlustering {#kplus}

@@

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

When defining the row vectors of $\mathbf{X'}_{N \times \mathit{TP}}$ with the notation $x'_{i.} = (x'_{i1}, \ldots, x'_{i(TP)})$ ($i = 1, \ldots, N$), the criterion $\mathit{SSE_{kplusGeneralized}}$ can be computed as

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

To thoroughly evaluate the k-plus approach to anticlustering, two specific k-plus criteria were compared to traditional anticlustering criteria in a second simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to group means and variances (by maximizing $\mathit{SSE_{kplus}}$); (ii) "generalized" k-plus anticlustering, minimizing differences with regard to group means, variances, skewness and kurtosis (by maximizing $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$). To implement the k-plus objectives, the unweighted standardization approach was employed that performed best in Simulation 1. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, (iii) a random allocation of elements to groups.

>>>>

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning. For likert scale data or binary attributes, the Manhattan distance may be preferred over the Euclidean distance, and for mixed-type attributes, it is even possible to employ a combined distance metric [@rokach2005].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, two simulation studies show that k-plus anticlustering is well-suited to create groups having minimal differences according to multiple distribution characteristics, significantly outperforming the traditional anticlustering criteria diversity and k-means. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to create groups that have similar variances across the $P$ attributes. As a caveat, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group mean $\overline{\mathbf{x}}^{(k)}$, but the additional variables computed to minimize differences with regard to the variance across groups are calculated using the overall mean $\overline{\mathbf{x}}$. Thus, $\mathit{SSE_{Var}}$ is only valid if (approximately) $\overline{\mathbf{x}}$ = $\overline{\mathbf{x}}^{(k)}$; otherwise we may introduce non-negligible measurement error when optimizing $\mathit{SSE_{Var}}$. Thus, only considering the new variables $\mathbf{Y}_{N \times P}$ in order to minimize discrepancies in variance may not yield satisfying results; instead, it has to be ensured that the group means of the original variables are also similar between groups. Fortunately, minimizing discrepancy with regard to group means can be accomplished by standard k-means anticlustering, i.e., the maximization of $\mathit{SSE}$. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

## A bicriterion model: K-plus anticlustering {#kplus}

@@

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *t-moment variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

When defining the rows of $\mathbf{X'}_{N \times \mathit{TP}}$ with the notation $x'_{i \cdot} = (x'_{i1}, \ldots, x'_{i(TP)})$ ($i = 1, \ldots, N$), the criterion $\mathit{SSE_{kplusGeneralized}}$ can be computed as

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

To thoroughly evaluate the k-plus approach to anticlustering, two specific k-plus criteria were compared to two traditional anticlustering criteria: (a) "standard" k-plus anticlustering, minimizing differences with regard to group means and variances (by maximizing $\mathit{SSE_{kplus}}$); (b) "generalized" k-plus anticlustering, minimizing differences with regard to group means, variances, skewness and kurtosis (by maximizing $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$). To implement the k-plus objectives, the unweighted standardization approach was employed that performed best in Simulation 1. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, (iii) a random allocation of elements to groups.
